---
title: "Robust estimators for Functional Partial Linear Models"
author: "Matias Salibian-Barrera"
date: "`r format(Sys.Date())`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy=TRUE)
```
This repository contains `R` code to compute the robust MM-estimators 
for Functional Partial Linear models in 
[Boente G, Salibian-Barrera M, Vena P. Robust estimation for semi-functional linear regression models. To appear in *Computational Statistics and Data Science*](https://github.com/msalibian/RobustFPLM). 
Original code by [Pablo Vena](https://github.com/anevolbap).

#### Installing the `RobustFPLM` package for `R`

The package can be installed directly from this repository using the 
following command in `R`:
```{r install, eval=FALSE}
devtools::install_github('msalibian/RobustFPLM', ref = "master")
```


### Tecator example
Because the regression model of the Tecator data includes an 
interaction between the non-parametric 
term `g` and a scalar covariate `w` (specifically, the model is
`y = <X, beta> + w * g(z) + e`, where `X` is the functional 
predictor, `w` and `z` are
scalar variables, and `g` is a smooth unknown function), setting 
the appropriate design matrices
(including splitting the data into a training and a testing set, 
and computing the 2nd derivatives
of the absorbance spectra) requires special care. 
Thus, we prepared a suite of specific 
functions and scripts to reproduce our analysis in the paper. 
See [here](tecator-specific-functions.R) and [here](script-TECATOR.R).

### A detailed example 

Here we will fit a FPLM to the [Tecator](http://lib.stat.cmu.edu/datasets/tecator) data set. 
It is also available from 
the [fda.usc](https://cran.r-project.org/package=fda.usc)
package.

The main function in our Robust FPLM suite of functions
is `FPLM-Bsplines()`. Its arguments include: 

- `y`: the vector of scalar responses;
- `x`: a matrix of  the functional covariates, 
where each row contains the functions evaluated 
on a (common) grid; 
- `t`: the grid over which 
the functional covariates were evaluated; 
- `u`: the values of the explanatory variable 
that enters the model non-parametrically;
- `range_freq`: a vector of B-spline basis sizes
to try for the functional regression coefficient;
- `range_spl`: a vector of B-spline basis sizes
to try for the non-parametric component;
- `norder`: the order of the B-splines;
- `fLoss`: the loss function to be minimized; and
- `trace`: a logical argument indicating whether
partial results are printed. 

We load our functions from the file `FPLM-Bsplines-functions.R`.
Note that to compute the robust estimators for the FPLM model we 
need the [robustbase](https://cran.r-project.org/package=robustbase)
package. 
```{r loadfunc, message=FALSE, warning=FALSE}
library(RobustFPLM)
```
Next, we load the `tecator` data 
from the `fda.usc` package and 
extract the response vector of fat content, the 
matrix of functional explanatory functional variables 
(the second derivatives of the absorption spectra),
the grid (t) on which the x's were evaluated,
and the vector (u) of protein values (the covariate that 
is modelled non-parametrically). 
The model is
`fat = < beta, absorp2> + eta(protein) + e`.
```{r loaddata, message=FALSE, warning=FALSE}
library(fda.usc)
data(tecator)
absorp2 <- fdata.deriv(tecator$absorp.fdata, nderiv = 2)
y  <- tecator$y$Fat[1:155]
u  <- tecator$y$Protein[1:155]
t  <- absorp2$argvals
x  <- absorp2$data[1:155,]
```
We consider B-splines of order 4, 
with bases sizes 
between 4 and 13, and use the Robust BIC criterion to 
determine the optimal B-splines dimensions
for each smooth component
```{r robustsetting}
norder <- 4
range_beta <- 4:13
range_eta <- 4:13
criterion <- 'bic1'
```
Now we fit the robust MM estimator:
```{r robust, tidy=TRUE, warning=FALSE, message=FALSE, cache=TRUE}
set.seed(124) # for lmrob()
a <- FPLMBsplines(y = y, x = x, u = u, t = t, range_freq = range_beta, range_spl = range_eta, norder = norder, fLoss='lmrob', trace=FALSE)
```
To fit the M and LS estimators we simply change the argument
`fLoss`:
```{r nonrobust, tidy=TRUE, warning=FALSE, message=FALSE, cache=TRUE}
d <- FPLMBsplines(y = y, x = x, u = u, t = t, range_freq = range_beta, range_spl = range_eta, norder = norder, fLoss='huang', trace=FALSE)

b <- FPLMBsplines(y = y, x = x, u = u, t = t, range_freq = range_beta, range_spl = range_eta, norder = norder, fLoss='ls', trace=FALSE)
```
The following plots compare the different fits. First 
we look at the non-parametric component:
```{r ploteta, fig.width=5, fig.height=5, tidy=TRUE}
u_order <- order(u)
ma <- max(rra <- c(a$fit$eta_est, b$fit$eta_est, d$fit$eta_est) )
mi <- min(rra)
plot(u[u_order], a$fit$eta_est[u_order], type = "l", xlab = "Protein", pch = 1, lwd = 6, col = "black", ylab = expression(hat(eta)), lty=1, ylim=c(mi, ma))
lines(u[u_order], d$fit$eta_est[u_order], lwd = 6, col = "tomato3")
lines(u[u_order], b$fit$eta_est[u_order], lwd = 6, col = "skyblue3")
legend('bottomright', lty=1, lwd=3, col=c('black', 'tomato3', 'skyblue3'), legend=c('MM', 'M', 'LS'))
```

And finally we look at the estimates for the functional 
regression coefficient:
```{r plotbeta, fig.width=5, fig.height=5, tidy=TRUE}
ma <- max(rrb <- c(a$fit$slope_fun, b$fit$slope_fun, d$fit$slope_fun) )
mi <- min(rrb)
plot(t, a$fit$slope_fun, col = "black", lwd = 6, type = "l", xlab = "Wavelength", ylab = expression(hat(beta)), lty=1, ylim=c(mi, ma))
lines(t, d$fit$slope_fun, col = "tomato3", lwd = 6, lty=1)
lines(t, b$fit$slope_fun, col = "skyblue3", lwd = 6, lty=1)
legend('bottom', lty=1, lwd=3, col=c('black', 'tomato3', 'skyblue3'), legend=c('MM', 'M', 'LS'))
```

We can now compare the fit obtained with the robust estimator to that 
obtained with the ``cleaned'' data, that is: we remove potential outliers
and fit the LS estimator to this ``clean'' data. 
To identify possibly atypical observations, we look at the residuals from the
robust fit, and flag those with large residuals (in absolute value):
```{r residuals, fig.width=5, fig.height=5, tidy=TRUE}
res <- y - a$fit$fitted
names(res) <- 1:length(res)
outl <- as.numeric(names(boxplot(res, col='gray70')$out))
```

We remove these suspect observations and re-fit the M- and LS-estimators
on the remaining ``clean'' data: 
```{r refit, cache=TRUE}
y2  <- y[ -outl ]
u2  <- u[ -outl ]
x2  <- x[ -outl,]
d2 <- FPLMBsplines(y = y2, x = x2, u = u2, t = t, range_freq = range_beta, range_spl = range_eta, norder = norder, fLoss='huang', trace=FALSE)

b2 <- FPLMBsplines(y = y2, x = x2, u = u2, t = t, range_freq = range_beta, range_spl = range_eta, norder = norder, fLoss='ls', trace=FALSE)
```
We now plot the robust estimator computed earlier on the whole data 
set, and the M- and LS-estimators re-computed on the ``cleaned'' data:
```{r ploteta2, fig.width=5, fig.height=5, tidy=TRUE}
u2_order <- order(u2)
ma <- max(rra <- c(a$fit$eta_est, b$fit$eta_est, d$fit$eta_est) )
mi <- min(rra)
plot(u[u_order], a$fit$eta_est[u_order], type = "l", xlab = "Protein", pch = 1, lwd = 6, col = "black", ylab = expression(hat(eta)), lty=1, ylim=c(mi, ma))
lines(u2[u2_order], d2$fit$eta_est[u2_order], lwd = 6, col = "tomato3")
lines(u2[u2_order], b2$fit$eta_est[u2_order], lwd = 3, col = "skyblue3")
legend('bottomright', lty=1, lwd=3, col=c('black', 'tomato3', 'skyblue3'), legend=c('MM', 'M2', 'LS2'))
```

and
```{r plotbeta2, fig.width=5, fig.height=5, tidy=TRUE}
ma <- max(rrb <- c(a$fit$slope_fun, b$fit$slope_fun, d$fit$slope_fun) )
mi <- min(rrb)
plot(t, a$fit$slope_fun, col = "black", lwd = 6, type = "l", xlab = "Wavelength", ylab = expression(hat(beta)), lty=1, ylim=c(mi, ma))
lines(t, d2$fit$slope_fun, col = "tomato3", lwd = 6, lty=1)
lines(t, b2$fit$slope_fun, col = "skyblue3", lwd = 3, lty=1)
legend('bottom', lty=1, lwd=3, col=c('black', 'tomato3', 'skyblue3'), legend=c('MM', 'M2', 'LS2'))
```

Note that all estimators are now essentially identical to each other. 
In other words, we see that the 
robust estimator results in a fit that is intuitively equivalent to using
LS on a ``cleaned'' data set. However, the robust procedure accomplishes this
automatically, and does not involve diagnostic plots or manually
removing observations from the data set. 

<!-- <!-- badges: start --> [![Travis build status](https://travis-ci.org/anevolbap/RobustFPLM.svg?branch=package)](https://travis-ci.org/anevolbap/RobustFPLM) <!-- badges: end --> -->